Depolarization of interacting NV$^-$ centers at low magnetic fields

### 2022-07-08 Mes papiers/Low field CR/V3/SI/Low_field_CR_SI.tex
la tete au weekend

\begin{document}
\vspace{0.2in}
{\Large \hspace{1.6in}\textsc{Supplementary Material} }\\
\
\title{Low magnetic field depolarization of NV$^-$ centers through dipole-dipole interaction}

\author{C. Pellet-Mary$^1$, M. Perdriat$^1$, P. Huillery$^?$,  G. H\'etet}

\affiliation{Laboratoire De Physique de l'\'Ecole Normale Sup\'erieure, \'Ecole Normale Sup\'erieure, PSL Research University, CNRS, Sorbonne Universit\'e, Universit\'e Paris Cit\'e , 24 rue Lhomond, 75231 Paris Cedex 05, France.}

@@

\section{NV$^-$ ground state Hamiltonian in low magnetic field}
\label{sec Hamiltonian}
\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_ESR}
\caption{ODMR measurement (a) in zero magnetic field, (b) in non-zero magnetic field when zooming on a single class}
\label{ESR_single_spin}
\end{figure}
With zero external magnetic field, there are three possible sources of splitting of the $\{\ket{+1},\ket{-1}\}$ manifold : local electric field, crystal strain and local magnetic field. Of these three causes, only the electric field can explain the shape of the ODMR line that we observe in zero external magnetic field (Fig. \ref{ESR_single_spin}a) \cite{mittiga2018imaging}.

Indeed, random local magnetic field would produce a single broadened line while crystal strain would produce a shifting of the zero field splitting (ZFS) of the same order as the splitting of the levels, which would also blur the two transitions into a single line. On the other hand, due to the large difference between the longitudinal and transverse electric field susceptibilities ($d_\parallel = 0.35\ \rm Hz\, cm/V$ and $d_\perp = 17\ \rm Hz\, cm/V$ \cite{van1990electric}), random local electric field will, on average, cause a splitting much stronger than its ZFS shifting and result in a two peak spectrum.

In our low magnetic field dipole-dipole coupling model, we will therefore neglect the contribution of the strain, local magnetic field and longitudinal electric field.  We also don't take into account the hyper-fine structure of the NV center due to the large inhomogeneous broadening of the transitions (Fig. \ref{ESR_single_spin}b).
We will then consider the following spin Hamiltonian for the NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_s=D S_z^2 + \gamma_e \bm{B}_{\rm ext} \cdot \bm{S}+ d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
Where $D=2.87\ \rm GHz$ is the zero field splitting and $\gamma_e=2.8\ \rm MHz/G$ the gyromagnetic ratio of the electron.

In the absence of an external magnetic field, the symmetry in the ($xy$) plane allows us to chose the $x$ direction along the electric field. The eigenstates of $\mathcal{H}_s$ then become $\{ \ket{0},\ket{+}=\frac{\ket{+1}+\ket{-1}}{\sqrt{2}},\ket{-}=\frac{\ket{+1}-\ket{-1}}{\sqrt{2}} \} $.

In the presence of non-zero magnetic field, let us call the Hamiltonian eigenstates $\{ \ket{g},\ket{d}, \ket{e} \} $. Fig. \ref{map etats propres} shows how close the $\ket{e}$ state is to the $\ket{+1}$ and $\ket{+}$ states as a function of the external magnetic field. Looking at the closeness of $\ket{d}$ to $\ket{-1}$ and $\ket{-}$ would show similar results, while $\ket{g}$ is pretty much equals to $\ket{0}$ for $B<100\ \rm G$.

This tells us that in most cases, the $\{ \ket{0},\ket{+},\ket{-} \}$ basis is only the good Hamiltonian basis for magnetic field smaller than a few Gauss, except in the case of pure transverse magnetic field where the $\{ \ket{0},\ket{+},\ket{-} \}$ basis remains a good basis even for sizable magnetic fields.
\begin{figure}
\includegraphics[width=0.7\textwidth]{Figures_SI/map_etats_propres}
\caption{Numerical simulations of the closeness of the Hamiltonian most excited state $\ket{e}$ with the states $\ket{+1}$ and $\ket{+}$, as a function of the magnetic field amplitude and angle $\theta$ with respect to the NV axis. A value of $d_\perp E_\perp = 4\ \rm MHz$ was chosen}
\label{map etats propres}
\end{figure}

\section{Samples}
%Note pour moi : j'ai inversé adamas 1 et 2 parraport aux noms des dossiers dans 20220324
Here are the various samples used in this study :
\begin{itemize}
\item \textbf{HPHT-150-1} : A 150 $\mu$m HPHT 1b diamond irradiated and annealed to reach a concentration [NV$^-$] $\approx 3\ \rm ppm$. Bought as is from Adamas Nanotechnology (MDNV150umHi). This sample is used in Fig. 2 (b), 2(d) and 3 of the main text, as well as Fig. \ref{ESR_single_spin}, \ref{T1_fits} and \ref{alphas} of the SI.
\item \textbf{HPHT-150-2} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. 4 of the main text and Fig. \ref{largeur_fluct} (c) of the SI.
\item \textbf{HPHT-150-3} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. \ref{Pola} of the SI.
\item \textbf{HPHT-15-1}  : A 15 $\mu$m diamond with similar properties as the last ones, also bought from Adamas Nanotechnology (MDNV15umHi). This sample is used in Fig. 5 of the main text.
\item \textbf{HPHT-15-2}  : Another sample from the same batch as HPHT-15-1. This sample is used in Fig. \ref{121 VS 22 fig} and \ref{Various ODMR} of the SI.
\item \textbf{CVD-PPM} : A CVD bulk diamond described in \citep{TALLAIRE2020421}. It has been irradiated and annealed and contains [NV$^-$] $\approx 4\ \rm ppm$. This sample is used in Fig. \ref{Alignment} of the SI.
\item \textbf{CVD-PPB} : Another CVD bulk diamond. This one has not been irradiated or annealed and therefore contains [NV$^-$] $\approx 50\ \rm ppb$. This sample is used in Fig. 2(a) of the main text.

@@

\section{Experimental Setup}
\begin{figure}
\includegraphics[width=0.7\textwidth]{Figures_SI/shema_exp}
\caption{Experimental Setup}
\label{setup}
\end{figure}
Fig. \ref{setup} shows the general purpose experimental setup used for all the experiments presented in this article.

@@

The laser used here is a Picoquant PDL 800-D with a 520 nm LDH laser head, providing pulses of 40 ps at a rate of 20 MHz, with an average power of $0.5 \sim 5\ \rm mW$. The trigger of the pulses is generated externally in order to achieve fast  gating of the laser for the $T_1$ measurement. We previously did similar experiment using a continuous 532 nm laser and observed no difference in the behavior of the spins.

The magnetic field is provided by a homemade electromagnet composed of a c-shape iron core and  copper wires. The magnet is mounted on two mechanical rotation stages, allowing a control on the polar and azimuthal angle of the magnetic field within a fraction of a degree and is alimented through a low frequency amplifier (Leybold power function generator 522 63)

The microwave field is generated by a Rhode \& Shwarz SMB 100A and is emitted with a handmade loop antenna. The field is gated by a Mini-circuits ZASWA-2-50DRA+ switch controlled externally and amplified by a Mini-circuits ZHL-5W-422+ amplifier.

@@

We did not use magnetic shielding to protect from the earth magnetic field since the splitting due to  $B_{\rm earth}$ is lower than the splitting due to the local electric field for the samples used here : $ \gamma_e B_{\rm earth} \approx 1.5\ \rm MHz < d_\perp E_\perp \approx 4\ \rm MHz$. 
When possible, the electromagnet was roughly aligned along the earth magnetic field to better compensate it.
\section{Experimental details}
\subsection{$T_1$ fitting Protocol}
The $T_1$ profile that we observe with our samples is neither purely exponential, nor purely stretched exponential, which is to be expected when the dipole-dipole spin decay is comparable to the phonon decay.

@@

An other possibility would be to use an arbitrary stretched factor in the fitting function : $S(\tau)=A \exp (-(\frac{\tau}{T_1^{\rm ph}})^\beta )$. Fig. \ref{alphas} shows the optimal $\beta$ parameter as a function of the external magnetic field, which confirms that the $T_1$ profile gets closer to a pure stretched-exponential in zero magnetic field.
\begin{figure}
\includegraphics[width=0.8\textwidth]{Figures_SI/Fig_T1_combined}
\caption{$T_1$ measurement with purely exponential and purely stretched exponential fits (a) in zero magnetic field (b) in non-zero magnetic field}
\label{T1_fits}
\end{figure}
\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures_SI/fig_alphas}
\caption{Best stretch factor $\beta$ for a $T_1$ fit of the form $f(\tau)=A \exp(-(\frac{\tau}{T_1})^\beta)$ as a function of a randomly oriented magnetic field amplitude}
\label{alphas}
\end{figure}
\subsection{Spectral range of the dipole-dipole cross-relaxations}
\label{fluctuator width}
\begin{figure}
\includegraphics[width=0.7\textwidth]{Figures_SI/largeur_fluct_SI}
\caption{Dipole-dipole depolarization for two near-resonant classes. (a) Sketch of the four possible orientations ("classes") in a single crystal diamond lattice. (b) ODMR spectrum showing four $\ket{0} \to \ket{-1}$ resonances corresponding to the four spin classes. The detuning $\Delta \nu$ between the classes b and c was controlled by changing the orientation of the external magnetic field.(c) Stretch part of the lifetime decay for the spins resonant with $\nu_c$ as a function of the detuning $\Delta \nu$ (blue circles), fitted by a Lorentzian with half width at half maximum 8.04 MHz. Single class ESR line stretched by a factor of 1,$\sqrt{2}$ and 2 are added for comparison}
\label{largeur_fluct}
\end{figure}

An experimental signature of the fluctuator hypothesis developed in \cite{choi_depolarization_2017} is to measure the depolarization rate for two near-resonant classes : if there are indeed very fast decaying NV centers (fluctuators with lifetime $T_1^f < 100 ns$), then the spectral width of the fluctuator would be increased beyond $1/T_2^*$ (which we assume to be homogeneous among all spins in the crystal), which means that they would be able to exchange spin quanta (flip-flop) with non-resonant NV detuned by $\Delta \nu$ such that $2\pi/T_1^f > \Delta \nu > 2\pi/T_2^*$

In order to verify this claim, we have to measure the spectral overlap between two classes, which in the absence of fluctuator should be proportional to the flip-flop rate, and compare it to the actual flip-flop rate which we can measure through the depolarization rate of the spins.

Fig. \ref{largeur_fluct} shows the results of such an experiment where we measured the stretched part of the NV's decay rate, which is very well fitted by a Lorentzian of width $\sigma=8.04 MHz$ (FW or HW ?), and compare it to an ODMR line of a single class of NV centers, stretched by a factor of $\sqrt{2}$ and 2 to simulate the spectral overlap of the two classes (discussed bellow). One should notice that, not only is $1/T_1^{\rm dd}$ significantly larger than the ODMR profile, it also doesn't have the same shape.

We should note that this broadening can not be explained by the dipole-dipole interaction strength : for a sample with 3 ppm of NV centers, the average dipole-dipole interaction strength between two nearest NV centers $J_0/r^3 \sim 27\ \rm kHz$ which is several order of magnitude lower than the broadening we observe.

\begin{figure}
\includegraphics[width=0.7\textwidth]{Figures_SI/overlap}
\caption{Illustration of the spectral overlap for two gaussian spectra}
\label{overlap}
\end{figure}

As illustrated on Fig. \ref{overlap}, we define the spectral overlap $S(\Delta \nu)$ between two spins of spectral response $S_1(\nu)$ and $S_2(\nu)$, centered respectively on the frequencies $\nu_1$ and $\nu_2$ where $\Delta \nu = \nu_2-\nu_1$ as :
\begin{equation}
S(\Delta \nu)=\int S_1(\nu, \nu_1)S_2(\nu, \nu_2) d\nu
\end{equation}
In order to approximate the spectral overlap in our experiment, we will consider the analytical solution in the Gaussian and Lorentzian case :

@@

\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures_SI/Fig_splitting}
\caption{(a) : Simulation of the 8 possible transition frequencies (four $\ket{0} \to \ket{-1}$ and four $\ket{0} \to \ket{+1}$) as a function of the magnetic field amplitude for the orientation of the magnetic field shown in Fig. 3(b) of the main text. (b) Difference in frequency between each pair of closest classes as a function of the magnetic field amplitude. The dotted line y=8.04 MHz correspond to the estimation of the dipole-dipole interaction range.}
\label{splitting}
\end{figure}
We give here an estimate of the magnetic field required to lift the degeneracy of the four classes in the case of Fig. 3(b) of the main text. Fig. \ref{splitting} shows the simulated energies for the 4 $\ket{0} \to \ket{-1}$ and the 4 $\ket{0} \to \ket{+1}$ as a function of the magnetic field, and the difference in energy between the closest pairs of transitions. We can see that the energy detuning between the classes $\Delta \nu$ crosses the value of the dipole-dipole interaction range (8.04 MHz, measured on sample HPHT-150-2 but likely to be similar on sample HPHT-150-1) for magnetic field values $\norm{B} = 12 \sim 15\ \rm G$, which are close to the half-width of the zero-field feature in Fig. 3(b)(ii) and (iii) of the main text.

@@

We did not observe a strong dependence on the laser polarization with the samples used in this study. Fig.\ref{Pola} shows the photoluminescence from one of our sample as a function of the magnetic field, either directly or with a modulation of the magnetic field, for 6 randomly chosen polarization angle and saw no major difference.

In particular, we did not observe the apparition of an anti-line inside the main dip, unlike what was observed in the two previously cited work. We expect that the main reason behind the different behaviors is that we seem to observe far grater spin depolarization in zero-field, which we attribute to dipole-dipole coupling, and that this effect could hide smaller effects such as the one related to the laser polarization.
\begin{figure}
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_Pola}
\caption{Effect of the polarization of the incident laser. (a) Photoluminescence of sample HPHT-150-3 as a function of randomly oriented magnetic field amplitude for various polarization angle. (b) Demodulated PL in the same conditions}
\label{Pola}
\end{figure}
\subsection{Alignment of the magnetic field along [100] axis}
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{Figures_SI/alignement}
\caption{PL of sample CVD-PPM as a function of the external magnetic field amplitude for different field misalignment with the [100] axis}
\label{Alignment}
\end{figure}
We claim that the decrease in PL at low magnetic field in Fig. 3 (c)(ii) of the main text is due to to the local electric field and the double flip processes. One other possibility would be a misalignment of the magnetic field with respect to the [100] axis, where the four classes would only be truly resonant with zero external magnetic field.

Fig. \ref{Alignment} shows the change in PL with respect to the external magnetic field amplitude for various misalignment of the magnetic field direction compared to the [100] diamond crystalline axis for sample CVD-PPM. We can see that the central drop in PL is not critically affected by the misalignment of the magnetic field, and we are therefore confident that the drop in PL we observe when $B\parallel [100]$ is not an effect of misalignment.

@@

Here are the results of the fluctuator model developed in \cite{choi_depolarization_2017} that we extended to include more scenarios. The following sections will detail the calculations that led to these results.
\begin{itemize}
\item The dipole-induced spin decay is at its lowest for a class of NV centers, non resonant with the three other classes, and dominated by a longitudinal magnetic field. In this case the dipole-dipole relaxation is limited by flip-flop processes between spins of the same class. We will call the corresponding stretched lifetime $T_1^{\rm dd}\equiv T_0$.
\item For a single class, still spectrally isolated   from the three other classes, but this time dominated by the electric field or transverse magnetic field (as long as $\gamma_e B_\perp \ll D$), the change in the eigenstates of the single particle Hamiltonian results in an increase of the average flip-flop rate. This leads to a new theoretical stretched lifetime $T_1^{\rm dd}= 4\, T_0$.
\item When a magnetic field is applied along the [100] crystalline axis (for $5\ \rm G < \abs{B} \ll D$), all four classes are resonant and dominated by the longitudinal part of the magnetic field. This results, in increase in the average flip-flop rate compared to the isolated class case, and the predicted stretched exponential lifetime is $T_1^{\rm dd}\approx 42.8\, T_0$.
\item In zero external magnetic field, all four classes are also resonant but we assume that the single particle Hamiltonian of each spin is dominated by a random local electric field. The averaging of flip-flop in this case yields a stretched lifetime $T_1^{\rm dd}\approx 51.4\, T_0$, which is about 20\% higher than the predicted lifetime for $B \parallel$ [100].
\end{itemize}

@@

The NV$^-$ centers in the crystal are divided between two categories : "normal" NV centers (simply called NV) which, in the absence of dipole-dipole coupling would have a phonon-limited $T_1$ ($T_1^{NV}\sim \rm ms$), and fluctuators who are NV centers with an additional, fast, depolarization mechanism, such that their lifetime $T_1^f < 100\ \rm ns$. Having such a short lifetime, the fluctuators are almost unpolarized by the green laser, making them invisible in standard optical NV measurement protocol ($T_1$, ODMR, etc.).

>>>>

\begin{document}
\vspace{0.2in}
{\Large \hspace{1.6in}\textsc{Supplementary Material} }\\
\
\title{Depolarization of interacting NV$^-$ centers at low magnetic fields}

\author{C. Pellet-Mary$^1$, M. Perdriat$^1$, P. Huillery$^2$,  G. H\'etet$^1$}

\affiliation{$^1$ Laboratoire De Physique de l'\'Ecole Normale Sup\'erieure, \'Ecole Normale Sup\'erieure, PSL Research University, CNRS, Sorbonne Universit\'e, Universit\'e Paris Cit\'e , 24 rue Lhomond, 75231 Paris Cedex 05, France \\ $^2$ Univ Rennes, INSA Rennes, CNRS, Institut FOTON - UMR 6082, F-35000 Rennes, France}

@@

\section{NV$^-$ ground state Hamiltonian under low magnetic field}
\label{sec Hamiltonian}
\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_ESR}
\caption{ODMR measurement (a) under zero magnetic field, (b) under a magnetic field of $~ 50 G$ when zooming on a single NV class}
\label{ESR_single_spin}
\end{figure}
Under zero external magnetic field, there are three possible processes that can cause a splitting of the $\{\ket{+1},\ket{-1}\}$ states : local electric fields, crystal strain and local magnetic fields. Of these three causes, only the electric field can explain the shape of the ODMR line that is observed under zero external magnetic field (Fig. \ref{ESR_single_spin}a) \cite{mittiga2018imaging}.

Indeed, random local magnetic field would produce a single broadened line while crystal strain would shift the zero field splitting (ZFS) by the same order of magnitude as the splitting between the levels, which would merge the two transitions onto a single line. Due to the large difference between the longitudinal and transverse electric field susceptibilities ($d_\parallel = 0.35\ \rm Hz\, cm/V$ and $d_\perp = 17\ \rm Hz\, cm/V$ \cite{van1990electric}), random local electric field can, on average, cause a splitting that is much stronger than the ZFS shift and result in a two peak spectrum centered on 2.87 GHz.

\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/map_etats_propres}
\caption{Numerical simulations showing the closeness of the Hamiltonian second excited state $\ket{e}$ with the states $\ket{+1}$ and $\ket{+}$, as a function of the magnetic field amplitude and angle $\theta$ with respect to the NV axis. A value of $d_\perp E_\perp = 4\ \rm MHz$ was chosen}
\label{map etats propres}
\end{figure}

In our model for  dipole-dipole coupling under small magnetic fields, we will therefore neglect the contribution of the strain, local magnetic fields and longitudinal electric fields.  We also do not take into account the hyper-fine structure of the NV center because of the large inhomogeneous broadening of the transitions in all our samples (Fig. \ref{ESR_single_spin}b).
We will then consider the following spin Hamiltonian for the NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_s/h=D S_z^2 + \gamma_e \bm{B}_{\rm ext} \cdot \bm{S}+ d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
Where $D=2.87\ \rm GHz$ is the zero field splitting and $\gamma_e=2.8\ \rm MHz/G$ the gyromagnetic ratio of the electron.

In the absence of an external magnetic field, the symmetry of the hamitlonian in the ($xy$) plane allows us to pick the $x$ direction along the electric field. The eigenstates of $\mathcal{H}_s$ then become $\{ \ket{0},\ket{+}=\frac{\ket{+1}+\ket{-1}}{\sqrt{2}},\ket{-}=\frac{\ket{+1}-\ket{-1}}{\sqrt{2}} \} $.

In the presence of a magnetic field at an angle $\theta$ with respect to the NV axis, we denote the Hamiltonian eigenstates $\{ \ket{g},\ket{d}, \ket{e} \} $ in ascending order of energy. Fig. \ref{map etats propres} shows how close the $\ket{e}$ state is to the $\ket{+1}$ and $\ket{+}$ states as a function of the external magnetic field and angle $\theta$. A similar analysis for the  $\ket{d}$, $\ket{-1}$ and $\ket{-}$ states show similar results, while $\ket{g}$ is pretty much equal to $\ket{0}$ for $B<100\ \rm G$.

This result tells us that, in most cases, the $\{ \ket{0},\ket{+},\ket{-} \}$ basis is only the good Hamiltonian basis for magnetic field smaller than a few Gauss, except in the case of pure transverse magnetic field where the $\{ \ket{0},\ket{+},\ket{-} \}$ basis remains a good basis even for sizable magnetic fields.

\section{Samples}
%Note pour moi : j'ai inversé adamas 1 et 2 parraport aux noms des dossiers dans 20220324
Here are the various samples used in this study :
\begin{itemize}
\item \textbf{HPHT-150-1} : A 150 $\mu$m HPHT 1b diamond irradiated and annealed to reach a concentration [NV$^-$] $\approx 3\ \rm ppm$. This sample was bought from Adamas Nanotechnology (MDNV150umHi). It is used in Fig. 2 (b), 2(d) and 3 of the main text, as well as Fig. \ref{ESR_single_spin}, \ref{T1_fits} and \ref{alphas} of the SI.
\item \textbf{HPHT-150-2} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. 4 of the main text and Fig. \ref{largeur_fluct} (c) of the SI.
\item \textbf{HPHT-150-3} : Another sample from the same batch as HPHT-150-1. This sample is used in Fig. \ref{Pola} of the SI.
\item \textbf{HPHT-15-1}  : A 15 $\mu$m diamond with similar properties as the last ones, also bought from Adamas Nanotechnology (MDNV15umHi). This sample is used in Fig. 5 of the main text.
\item \textbf{HPHT-15-2}  : Another sample from the same batch as HPHT-15-1. This sample is used in Fig. \ref{121 VS 22 fig} and \ref{Various ODMR} of the SI.
\item \textbf{HPHT-1-1}  : A 1 $\mu$m diamond with similar properties as the last ones, also bought from Adamas Nanotechnology (MDNV1umHi). This sample is used to perform a sensitivity measurement similar to the one done in Fig. 5 of the main text.
\item \textbf{CVD-PPM} : A CVD bulk diamond described in \citep{TALLAIRE2020421}. It has been irradiated and annealed and contains [NV$^-$] $\approx 4\ \rm ppm$. This sample is used in Fig. \ref{Alignment} of the SI.
\item \textbf{CVD-PPB} : Another CVD bulk diamond. This one has not been irradiated or annealed and therefore contains [NV$^-$] $\approx 50\ \rm ppb$. This sample is used in Fig. 2(a) of the main text.

@@

\section{Experimental Setup}
\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/shema_exp}
\caption{Experimental Setup}
\label{setup}
\end{figure}
Fig. \ref{setup} shows the general purpose experimental setup used for all the experiments presented in this article.

@@

The laser used here is a Picoquant PDL 800-D with a 520 nm LDH laser head, providing pulses of 40 ps at a rate of 20 MHz, with an average power of $0.5 \sim 5\ \rm mW$. These fast pulses serve no purpose in the actual experiment. This laser was used solely because of its high amplitude stability. The trigger of the pulses is generated externally in order to achieve fast gating of the laser for the $T_1$ measurement. We previously did similar experiment using a continuous 532 nm laser and observed no difference in the spin response.

The magnetic field is provided by a homemade electromagnet composed of a c-shape iron core and  copper wires. The magnet is mounted on two mechanical rotation stages, allowing a control on the polar and azimuthal angle of the magnetic field within a fraction of a degree and is alimented through a low frequency amplifier (Leybold power function generator 522 63).

The microwave field is generated by a Rhode \& Shwarz SMB 100A and is emitted with a handmade loop antenna. The microwave field is gated by a Mini-circuits ZASWA-2-50DRA+ switch controlled externally and amplified by a Mini-circuits ZHL-5W-422+ amplifier.

@@

We did not use magnetic shielding to protect from the earth magnetic field since the splitting due to  $B_{\rm earth}$ is lower than the splitting due to the local electric field for the samples used here : $ \gamma_e B_{\rm earth} \approx 1.5\ \rm MHz < d_\perp E_\perp \approx 4\ \rm MHz$. 
When possible, the electromagnet was roughly aligned along the earth magnetic field to better compensate it.
\section{Experimental details}
\subsection{$T_1$ fitting Protocol}
\begin{figure}
\includegraphics[width=0.8\textwidth]{Figures_SI/Fig_T1_combined}
\caption{$T_1$ measurement with purely exponential and purely stretched exponential fits (a) in zero magnetic field (b) in non-zero magnetic field}
\label{T1_fits}
\end{figure}
\begin{figure}[h]
\includegraphics[width=0.45\textwidth]{Figures_SI/fig_alphas}
\caption{Best stretch factor $\beta$ for a $T_1$ fit of the form $f(\tau)=A \exp(-(\frac{\tau}{T_1})^\beta)$ as a function of an arbitrarely oriented magnetic field amplitude}
\label{alphas}
\end{figure}

The $T_1$ profile that we observe with our samples is neither purely exponential, nor purely stretched exponential, which is to be expected when the dipole-dipole induced spin decay is comparable to the phonon induced decay.

@@

An other possibility would be to use an arbitrary stretched factor in the fitting function : $S(\tau)=A \exp (-(\frac{\tau}{T_1^{\rm ph}})^\beta )$. Fig. \ref{alphas} shows the optimal $\beta$ parameter as a function of the external magnetic field, which confirms that the $T_1$ profile gets closer to a pure stretched-exponential in zero magnetic field.

\subsection{Spectral range of the dipole-dipole cross-relaxations}
\label{fluctuator width}
\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/largeur_fluct_SI}
\caption{Dipole-dipole depolarization for two near-resonant classes. (a) Sketch of the four possible NV orientations (``classes") in a single diamond. (b) ODMR spectrum showing four $\ket{0} \to \ket{-1}$ resonances corresponding to the four spin classes. The detuning $\Delta \nu$ between the classes b and c was controlled by changing the orientation of the external magnetic field. (c) Stretched part of the lifetime decay curve for the spins resonant with $\nu_c$ as a function of the detuning $\Delta \nu$ (blue circles), fitted by a Lorentzian with half width at half maximum 8.04 MHz. Single class ESR line stretched by a factor of 1,$\sqrt{2}$ and 2 are added for comparison.}
\label{largeur_fluct}
\end{figure}

An experimental signature of the fluctuator hypothesis developed in \cite{choi_depolarization_2017} is the dependence of the depolarization rate when two near-resonant classes are brought to resonance : if there are indeed very fast decaying NV centers (fluctuators with lifetime $T_1^f < 100$ ns), then the spectral width of the fluctuator would be greater than $1/T_2^*$.  For simplicity, it is typically assumed that the latter is the same for all spins in the crystal. This means that NV centers would be able to exchange spin quanta (flip-flop) with non-resonant NV fluctuators detuned by $\Delta \nu$ such that $2\pi/T_1^f > \Delta \nu > 2\pi/T_2^*$.

In order to verify this, we measure the spectral overlap between two classes, which in the absence of fluctuator should be on the order of $1/T_2^*$, and compare it to the actual width of the depolarization rate of the spins
as a function of detuning.

Fig. \ref{largeur_fluct} shows the results of such an experiment. We measured the stretched part of the NV's decay rate at each detuning and obtained Fig. \ref{largeur_fluct} -c), which is very well fitted by a Lorentzian of half-width $\sigma=8.04$ MHz. 
We should note that this broadening can not be explained by the direct coherent dipole-dipole interaction induced splitting : for a sample with 3 ppm of NV centers, the average dipole-dipole interaction strength between two nearest NV centers $J_0/r^3 \sim 27\ \rm kHz$ which is several order of magnitude lower than the broadening we observe.

\begin{figure}[h]
\includegraphics[width=0.7\textwidth]{Figures_SI/overlap}
\caption{Illustration of the spectral overlap for two gaussian spectra}
\label{overlap}
\end{figure}

We can compare this obtained width to the width of an ODMR line of a single class of NV centers, stretched by a factor of $\sqrt{2}$ and 2 to simulate the spectral overlap of two classes. One should notice that, not only is $1/T_1^{\rm dd}$ significantly larger than the ODMR profile, it also does not have the same shape.

As illustrated in Fig. \ref{overlap}, we define the spectral overlap $S(\Delta \nu)$ between two spins of spectral response $S_1(\nu)$ and $S_2(\nu)$, centered respectively on the frequencies $\nu_1$ and $\nu_2$ where $\Delta \nu = \nu_2-\nu_1$ as :
\begin{equation}
S(\Delta \nu)=\int S_1(\nu, \nu_1)S_2(\nu, \nu_2) d\nu
\end{equation}
In order to approximate the spectral overlap in our experiment, we will consider the analytical solution in the Gaussian and Lorentzian case :

@@

\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figures_SI/Fig_splitting}
\caption{(a) : Simulation of the 8 possible transition frequencies (four $\ket{0} \to \ket{-1}$ and four $\ket{0} \to \ket{+1}$ transitions) as a function of the magnetic field amplitude for the orientation of the magnetic field shown in Fig. 3(b) of the main text. (b) Difference in frequency between each pair of closest classes as a function of the magnetic field amplitude. The dotted line y=8.04 MHz corresponds to an estimation of the dipole-dipole interaction frequency range.}
\label{splitting}
\end{figure}
We give here an estimate of the magnetic field required to lift the degeneracy of the four classes in the case of Fig. 3(b) of the main text. Fig. \ref{splitting} shows the simulated energies for the 4 $\ket{0} \to \ket{-1}$ and the 4 $\ket{0} \to \ket{+1}$ transitions as a function of the magnetic field, and the difference in energy between the closest pairs of transitions. We can see that the energy detuning between the classes $\Delta \nu$ crosses the value of the dipole-dipole interaction range (8.04 MHz, measured on sample HPHT-150-2 but likely to be similar on sample HPHT-150-1) for magnetic field values $\norm{B} = 12 \sim 15\ \rm G$, which are close to the half-width of the zero-field feature in Fig. 3(b)(ii) and (iii) of the main text.

@@

We did not observe a strong dependence on the laser polarization with the samples used in this study. Fig.\ref{Pola} shows the photoluminescence from one of our sample as a function of the magnetic field, either directly or with a modulation of the magnetic field, for 6 polarization angles and saw no major difference.

In particular, we did not observe the apparition of an anti-line inside the main dip, unlike what was observed in the two previously cited work. We expect that the main reason behind the different behaviors is that we seem to observe far greater spin depolarization in zero-field, which we attribute to dipole-dipole coupling. This effect could hide smaller effects such as the one related to the laser polarization.
\begin{figure}[h]
\includegraphics[width=0.9\textwidth]{Figures_SI/fig_Pola}
\caption{Effect of the polarization of the incident laser. (a) Photoluminescence of sample HPHT-150-3 as a function of arbitrarely oriented magnetic field amplitude for various polarization angle. (b) Demodulated PL in the same conditions.}
\label{Pola}
\end{figure}
\subsection{Alignment of the magnetic field along [100] axis}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{Figures_SI/alignement}
\caption{PL of sample CVD-PPM as a function of the external magnetic field amplitude for different field misalignment with the [100] axis.}
\label{Alignment}
\end{figure}
We claim that the decrease in PL at low magnetic field in Fig. 3 (c)(ii) of the main text is due to the local electric field and the double flip processes. One other possibility would be a misalignment of the magnetic field with respect to the [100] axis as it is scanned. The four classes could for instance be truly resonant only with zero external magnetic field and not when the B field is large.

Fig. \ref{Alignment} shows the change in PL with respect to the external magnetic field amplitude for various misalignments of the magnetic field direction compared to the [100] diamond crystalline axis of the sample CVD-PPM. We can see that the central drop in PL is not critically affected by the misalignment of the magnetic field, and we are therefore confident that the drop in PL we observe when $B\parallel [100]$ is not an effect of misalignment.

@@

Here are the results of the fluctuator model developed in \cite{choi_depolarization_2017} that we extended to include double flip as well as local electric fields. The following sections will detail the calculations that led to these results. Let us first summarize our theoretical findings :

\begin{itemize}
\item The dipole-induced spin decay is lowest for a spectrally isolated class of NV centers when the transverse magnetic field is negligible. In this case the dipole-dipole relaxation is limited by flip-flop processes between spins of the same class. We will call the corresponding stretched lifetime $T_1^{\rm dd}\equiv T_0$.
\item For a single class, still spectrally isolated   from the three other classes, but this time dominated by the electric field or transverse magnetic field (as long as $\gamma_e B_\perp \ll D$), the change in the eigenstates of the single particle Hamiltonian results in an increase of the average flip-flop rate. This leads to a new theoretical stretched lifetime $T_1^{\rm dd}= 4\, T_0$.
\item When a magnetic field is applied along the [100] crystalline axis (for $5\ \rm G < \abs{B} \ll D$), all four classes are resonant and dominated by the longitudinal part of the magnetic field. This results, in increase in the average flip-flop rate compared to the isolated class case, and the predicted stretched exponential lifetime is $T_1^{\rm dd}\approx 42.8\, T_0$.
\item In zero external magnetic field, all four classes are also resonant but we assume that the single particle Hamiltonian of each spin is dominated by a random local electric field. The averaging of flip-flop in this case yields a stretched lifetime $T_1^{\rm dd}\approx 51.4\, T_0$, which is about 20\% higher than the predicted lifetime for $B \parallel$ [100].
\end{itemize}

@@

The NV$^-$ centers in the crystal are divided between two categories :``normal" NV centers (simply called NV) which, in the absence of dipole-dipole coupling would have a phonon-limited $T_1$ ($T_1^{NV}\sim \rm ms$), and fluctuators who are NV centers with an additional, fast, depolarization mechanism, such that their lifetime $T_1^f < 100\ \rm ns$. Having such a short lifetime, the fluctuators are almost unpolarized by the green laser, making them invisible in standard optical NV measurement protocol ($T_1$, ODMR, etc.).
